---
title: "Class 8 : Breast Cancer Mini Project"
author: "Alfred Phyo (PID: A18129346)"
format: pdf
toc: TRUE
---

## Background

In today's class we will applu the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set.


## Data Import

We start by importing out data. It is a csv file so we will use the `read.csv()` function

```{r}
wisc.df <- read.csv("wisconsincancer.csv", row.names = 1)
```

have a wee peak at the first few rows:

```{r}
head(wisc.df)
```

Make sure to remove the first `diagnosis` column - I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)

length(grep("_mean", colnames(wisc.data)) )
```

## Principal Component Analysis

The main function here is `prcomp()` and we want to make sure we set the optional argument `scale=TRUE`:

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```
> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27% of the original variance is captured by the first principal component (PC1).

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Three principal components are required to describe 72.64% of the original variance.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Seven principal components are required to describe 91.01% of the original variance.

**Biplot:**

```{r}
biplot(wisc.pr)
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

There's too much information on the plot making it very difficult to understand because the labels and points are all over the place.


Our main PCA "score plot" or "PC plot" of results
```{r}
library(ggplot2)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2, color = diagnosis) +
  geom_point()
```
> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, color = diagnosis) +
  geom_point()
```

These plots are much more organized and easier to understand, allowing us to differentiate between malignant and benign.

Scree-plot:

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)

```


```{r}
ggplot(wisc.pr$rotation) +
  aes(PC1,
      reorder(rownames(wisc.pr$rotation), -PC1)) + 
        geom_col()
```

Collectively these two plot ("score plot" and "loadings plot") tell us that if cells nucleus are deeply indented ("concave"), irregular non circular ("compactness"), and have large "perimeter" values they tend to be malignant....

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```
The loading of concave.points_mean on PC1 is −0.2609, indicating a strong contribution to the first PC. Yes, there are several features have larger contributions, including concavity_mean, concave.points_worst, and size-related features such as perimeter_worst and radius_worst.

## Hierarchical clustering

First scale the data (with the `scale()` function), then calculate a distance matrix (with the `dist()` function). Then cluster with the `hclust()` function and plot:

```{r}
wisc.hclust <- hclust(dist(scale(wisc.data)))
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=20, col="red")
```
It has 4 clusters when height is at 20.

You can also use `cutree()` function with a argument `k=4` rather than `h=height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I like the ward.D2 method because it creates groups such that variance is minimized within clusters.


## Combining methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables) and use a subset of these PCs that capture the most variance as input for `hclust()`

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```
> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

The newly created hclust model separates out the two "M" and "B" diagnoses really well because there are now two distinct big groups in the cluster dendrogram.


Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

The hierarchical clustering done without PCA somewhat separates benign and malignant case, but there is still a lot of overlap between the clusters. When clustering is done after PCA, the separation is much clearer, with one cluster mostly malignant and the other mostly benign. Overall, using PCA before hierarchical clustering gives better separation of the diagnoses.

I want to know how the clustering in `grps` with values of 1 or 2 correspond to the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

My clustering **groups** 1 are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis (333)

24 FP
179 TP
333 TN
33 FN

Sensitivity TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333+24)
```


## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col = grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
> Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 2 because they are likely be be diagnosed as malignant compared to patient 1. This is because patient 2 falls inside the group with the most malignant cases.