---
title: "Class 12 HW: Population analysis"
author: "Alfred Phyo (PID: A18129346)"
format: pdf
---

## Section 4: Population analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
 
The sample size for each genotype are as follows: 

```{r}
table(expr$geno)
```

```{r}
AA <- expr$exp[expr$geno == "A/A"]
AG <- expr$exp[expr$geno == "A/G"]
GG <- expr$exp[expr$geno == "G/G"]

summary(AA)
summary(AG)
summary(GG)
```
The median for genotype **AA** is 31.25.
The median for genotype **AG** is 25.07.
The median for genotype **GG** is 20.07.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch=TRUE)
```

From the relative expression value between A/A and G/G displayed in this plot, I can infer individuals with A/A genotype have a higher expression of ORMDL3 than individuals with G/G genotype. The SNP appears to influence gene expression of ORMDL3, suggesting its contribution to asthma susceptibility.

