---
title: "Class 18: Pertussis Mini-Project"
author: "Alfred Phyo (PID: A18129346)"
format: pdf
toc: TRUE
---

# Background
Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year of age)

# CDC tracking data
The CDC tracks the number of Pertussis cases: 

```{r, echo = FALSE}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L, 2024L, 2025L),
                                cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538, 21996)
       )
```


> Q1. I want a plot of year vs cases

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```

> Q2. Add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996). What do you notice?

There is a big decrease in Pertussis cases from 1946 to 1996 after switching to the aP vaccine.

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", lty =2)+
  geom_vline(xintercept = 1996, col = "red", lty =2) + 
  geom_vline(xintercept = 2020, col = "gray", lty = 2)
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is an upward trend in Pertussis cases after the introduction of the aP vaccine in 1996. Some possible explanations are that less people are getting vaccinated, the bacteria could be evolving, and our methods on testing for these data are becoming more accurate than before.

> Additional points for discussion: How are vaccines currently approved?

Vaccines are currently approved by the FDA through rigorous clinical trials and evaluation of safety, efficacy, and manufacturing quality. It is more strict than how it was in the past.

## Exploring CMI-PB data

The CMI-PB project <https://www.cmi-pb.org/>
The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection.

They make their data availale as JSON format API. We can read this into R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> Q4. How many aP and wP individuals are there in this `subject` table?

```{r}
table(subject$infancy_vac)
```

> Q5. How many male/female are there?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Is this representative of the US population?

No, it's not representative of the US population.

We can read more tables from the CMI-PB database 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

To make sense of all this data we need to "join" (a.k.a. "merge" or "link") all these tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain date (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).

We can use **dplyr** and the `*join()` family of functions to do this. 

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta) 
```
> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata) 
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q. How many different "antigen" values are measured?

16 different antigen values are measured.

```{r}
table(abdata$antigen)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

The most "recent" dataset does not follow the downward trend. It seems to have a jump (doubled) in data compared to the previous year.

```{r}
table(abdata$dataset)
```

## Examine IgG Ab titer levels

Let's focus on IgG isotype

```{r}
igg <- abdata |>
        filter(isotype=="IgG")

head(igg)
```
> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

Make a plot of `MFI_normalised` values for all `antigen` values.

```{r}
library(ggplot2)

ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()+
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens "PT", "FIM2/3" and "FHA" appear to have the widest range of values. These antigens show differences in the level of IgG antibody titers recognizing them because they play a key role in the immune response (both for protection and diagnosis) for Bordetella pertussis.

> Q. Is there a difference for these responses between aP and wP individuals?

Yes.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
  facet_wrap(~infancy_vac)
```

> Q. Is there a difference with time (i.e. before booster shot vs. after booster shot)?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

```{r}
## filter to 2021 dataset, IgG and PT
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
  
# Individual subject trajectories (thin, faint)
    geom_point(alpha = 0.4, size = 1.5) +
    geom_line(alpha = 0.25, linewidth = 0.6) +
  
# Smooth bold average trendlines for aP and wP
  stat_summary(fun = mean,
               geom = "smooth",
               se = FALSE,
               linewidth = 1.6,
               span = 0.5,
               aes(group = infancy_vac, col = infancy_vac)) +
  
  # Reference lines
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="CMI-PB 2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and day 14 (post boost)")
```

> Q18. Does this trend look similar for the 2020 dataset?

No, it looks completely different since the 2021 dataset had a significant decrease from the 2020 dataset.

```{r}
## filter to 2020 dataset, IgG and PT only
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```