---
title: "HW6"
author: "Alfred Phyo (PID:A18129346)"
format: pdf
toc: TRUE
---

## Original Code

```{r}
# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

## Answer

> Q6. How would you generalize the original code above to work with any set of input protein structures?

```{r}
library(bio3d)
```

```{r}
plot_protein <- function(x){
ans <- read.pdb(x)
ans.chainA <- trim.pdb(ans, chain="A", elety="CA")

ans.b <- ans.chainA$atom$b

plotb3(ans.b, sse=ans.chainA, typ="l", ylab="Bfactor")
}
```

## Example output


```{r}
plot_protein("4AKE") #The input to the function are character strings specifying the PDB ID (e.g. 4AKE, 1AKE)
```

## Documentation (narrative text)

The function `plot_protein()` generates a B-factor plot for the user specified protein. 

The function can be used by the user when they input a character string specifying the PDB ID available in the database.

The function's output is a B-factor plot that demonstrates how much a protein moves or fluctuates in a crystal structure. 
